Contracts for unions and intersections are hard for Nickel

### 2021-06-04 paper.tex
Continue reworking section 4 (WIP)

\subsection*{Union contracts are hard to break}

@@

There is an intuitive justification for this fact. Apart from the error
reporting part (which is crucial in practice!), the current contract system of
Nickel can be implemented purely as a library, just given a \nickel{fail}
primitive to abort the execution. In particular, applying a function contract to
\nickel{f} replaces it with a guarding function \nickel{fWrapped} that performs
the required checks. Thus, in a pure setting, the failure of \nickel{fWrapped 5}
must be independant of its environment, and in particular of the call
\nickel{fWrapped 10}.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f | Positive -> Positive
      = fun x => x - 7 in
(f 10) + (f 5)
\end{lstlisting}
\caption{Wrong positive to positive function}
\label{fig:pos-to-pos}
\end{figure}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f | (Positive -> Positive)
        @| (Positive -> NonPositive)
      = fun x => x - 7 in
(f 10) + (f 5)
\end{lstlisting}
\caption{Still wrong positive to positive or positive to non-positive function}
\label{fig:wrong-union-function}
\end{figure}

In a system with union and intersection, this contract would fail, because
\nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
But, as opposed to the previous example, removing one call or the other will make the program pass!
Indeed, the first call only fails the \nickel{Positive -> NonPositive} part of
the union,
while the second one only fails the \nickel{Positive -> NonPositive} part. Both
are needed to prove the violation of the full contract.

Even in a pure setting, unions introduce a form of effect at a distance: the
result of \nickel{f 5} is not local anymore as it depends on the previous
execution of other calls to \nickel{f}. With unions,
equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations} does not
hold, and the candidate example of
Figure~\ref{fig:optimizations-let-floating-ex} cannot be optimized in general.

Figure~\ref{fig:optimized-programs} below shows another example of the same
fact. A seemingly valid optimization ends up with a program that behave
differently from the original one.

@@

This seemingly simple consequence generates major issues when considered in real
life languages.
For example, remember the different mechanisms for code manipulation presented
in Section \ref{sec:optimizations}, inlining and commmon subexpression elimination,
both of these rely on the fact that substituting a pure function call by its body,
or joining equal pure expressions and sharing their result, is a semantically valid
transformation.
However, consider the examples presented at Figure \ref{fig:optimized-programs},
the first one is an inlined version of the second one (where function
\nickel{g} has been inlined); while the second is similar to the first, but
the common subexpression \nickel{f 1} has been eliminated.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Inlined]
let f | Num -> (Bool -> Num @| Bool -> Str)
      = fun x y => if y then x else "False"
in (f 1 true, f 1 false)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
let f = (fun x y => if y then x else "False") | Num -> (Bool -> Num @| Bool -> Str) in
let g = f 1 in
(g true, g false)
\end{lstlisting}
\caption{Equivalent programs, with inlining or CSE applied}
\label{fig:optimized-programs}
\end{figure}

Unexpectedly, these two programs behave very differently, the first one will
return \nickel{(1, "False")}, without failing.  While the second one will fail,
since the single application of \nickel{f 1}, applied to both \nickel{true} and
\nickel{false} will be able to find the error on the union contract
\nickel{(Bool -> Num) @| (Bool -> Str)}.

>>>>

\subsection*{Union contracts as an effect}

@@

There is an intuitive justification for this fact. Apart from the error
reporting part (but this is the crucial in practice!), the current contract
system of Nickel can be implemented purely as a library, just given a
\nickel{fail} primitive to abort the execution. In particular, applying a
function contract to \nickel{f} replaces it with a guarding function
\nickel{fWrapped} that performs the required checks. Thus, in a pure setting,
the failure of \nickel{fWrapped 5} must be independant of its environment, and
in particular of the call \nickel{fWrapped 10}.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f | Positive -> Positive
      = fun x => x - 7 in
(f 10) + (f 5)
\end{lstlisting}
\caption{Simple contract violation}
\label{fig:pos-to-pos}
\end{figure}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f | (Positive -> Positive)
        @| (Positive -> NonPositive)
      = fun x => x - 7 in
(f 10) + (f 5)
\end{lstlisting}
\caption{Union contract violation}
\label{fig:wrong-union-function}
\end{figure}

In a system with unions, this contract would fail, because \nickel{f 10}
(returns \nickel{3}) is a witness of \nickel{f} failing the contract
\nickel{Positive -> NonPositive}, and \nickel{f 5} (returns \nickel{-2}) is a
witness of \nickel{f} failing \nickel{Positive -> Positive}.  But, as opposed to
the previous example, removing one call or the other will make the program pass!
Indeed, the first call only fails the \nickel{Positive -> NonPositive} part of
the union, while the second one only fails the \nickel{Positive -> NonPositive}
part. Both are needed to prove the violation of the full contract.

Even in a pure setting, unions introduce a form of effect at a distance: the
result of \nickel{f 5} is not local anymore, as it now depends on the previous
execution of other calls to \nickel{f}. Unions contracts break the
equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations}, and the
candidate example of Figure~\ref{fig:optimizations-let-floating-ex} cannot be
optimized in general.

Figure~\ref{fig:optimized-programs} illustrates this point further. It comprises
an original program, and one where the common subexpression \nickel{f 1} have
been eliminated. These two programs behave very differently:
\begin{itemize}
    \item The original version returns \nickel{(1, "False")} without failing.
    \item The optimized version fails with a contract violation.
\end{itemize}

since the single application of \nickel{f 1}, applied to both \nickel{true} and
\nickel{false} will be able to find the error on the union contract
\nickel{(Bool -> Num) @| (Bool -> Str)}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
let f | Num -> (Bool -> Num @| Bool -> Str)
      = fun x y => if y then x else "False"
in (f 1 true, f 1 false)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Optimized]
let f | Num -> (Bool -> Num @| Bool -> Str)
      = fun x y => if y then x else "False"
let g = f 1 in
(g true, g false)
\end{lstlisting}
\caption{Equivalent programs, with inlining or CSE applied}
\label{fig:optimized-programs}
\end{figure}

@@

To sum up, union contracts introduce a form of effect at a distance. This
impacts both the programmer and the compiler or interpreter:
\begin{itemize}
    \item As for any side-effect, this breaks local reasoning.
    \item It prevents optimizations.
\end{itemize}
\todo{WIP}
